Estimating real-time high street footfall from Wi-Fi probe requests

### 2018-01-30 abstract.tex
second draft

\title{Estimating real-time highstreet footfall from the Wi-Fi probe requests}
\author{
	\authorstyle{
		Balamurugan Soundararaj\textsuperscript{1}, 
		James Cheshire\textsuperscript{1} and 
		Paul Longley\textsuperscript{1}}
	\newline\newline
	\textsuperscript{1}\institution{
		Department of Geography, 
		University College London, 
		United Kingdom}
}
\date{\today}

@@

	% ---------------------------------------------------------------------------
	% We talk about wifi and probe requests. Possibility of 
	% counting people without revealing identity. There is uncertainity. We look
	% at resolving these. We have a setup to collect data from sensor and manual
	% We propose methodology to clean sensor count. compare with manual to see
	% it corresponds properly.
	% ---------------------------------------------------------------------------

@@

	\textbf{Abstract}

	Wi-Fi has become an ubiquitous technology used in provinding internet access in
	public and private spaces to people's mobile devices such as smartphones, 
	tablets and laptops. This has resulted in a situation
	where any point in a dense urban built environment 
	has multiple Wi-Fi networks available. Modern mobile devices anticipate and 
	remember these networks and to be able to switch seamlessly between them,
	broadcast a special type of management packets known as probe requests.
	This is a low level signal (probe request) specified in IEEE 802.11b/g specification
	which relays information about the source mobile device to
	any Access Points (AP) listening around them.
	This is the first step in establishing a connection between these devices and
	is universal to any device which has a Wi-Fi radio to communicate.
	This makes these probe requests a continuous, passive and wireless source of
	data available at any urban location which can be a clue in
	understanding the number of people present in that area in real-time (\citep{freud2015},\citep{konto2017}).
	In this paper, from a set of probe requests collected
	at a highstreet location in London, along with manually collected data, we demonstrate
	that pedestrian footfall can be estimated with reasonable accuracy without
	infringing on the privacy of the mobile users.

	\begin{figure*}
		\begin{center}
	  \includegraphics[width=0.8\textwidth]{outputs/clustering_1.png}
	  \caption{Clustering probe requests based on increasing sequence numbers generated over time.}
		\end{center}
	\end{figure*}

	There have been numerous attempts at using Wi-Fi
	to measure the volume and movement of people in built environment for various
	applications (\citep{zarim2006},\citep{sap2015},\citep{reki2007}).
	Though favorable results have been observed, one of the major 
	challanges been the changes in the MAC address randomisation process which 
	aims to protect the users' privacy by anonymising the only globally identifiable
	portion of the probe requests \citep{green2008}. There have been various 
	successful research in breaking this
	randomisation process to extract real MAC addresses \citep{martin2017}
	but this usually results in serious infringement on the user's
	privacy. There is a clear gap in the research for
	exploring methodologies with which we can estimate the
	number of unique mobile devices at a location without the need to fingerprint
	them using their MAC address which we try to address in this paper.

	The pilot survey was conducted on Oxford st. in London on 20 December 2017 from 12:30 to
	13:00. The data was collected parallely through two methods - Wi-Fi sensor and Android app.
	The Wi-Fi sensor collected all the probe requests broadcated around the area
	and recorded the timestamp at which they were colleted, MAC address of the source
	device anonymised using a hashing algorithm, organisationally unique identifier (OUI) of the
	vendor who manufactured the device, total length of the singal in bits, signal
	strength reported by the device in dBm, duration for which the signal was tranmmitted, 
	the service set identifier (SSID) to which the probe request has been sent to and the
	length of the extra tags in the packets. The android app recorded the timestamp
	of every touch on the screen by the surveyor which corresponds to a pedestrian 
	crossing a cordon on the sidewalk.

	Approximately 60,000 probe requests were collected out of which   \% had
	randomised MAC addresses thus effectively anonymised.
	When we include these anonymised probe requests and count the number of unique
	devices present in dataset, we find that the difference between manual counting and sensor based counting
	is 425\% on average thus showing the need for using a different fingerprint other than MAC address for the
	anonymised probe requests. We first filter the data by removing all the probe requests reporting
	low signal strength. This classification is done using a kmeans - class interval algorithm.
	This reduces the difference from the manual count to 30\%.

>>>>

\title{Estimating real-time high street footfall from Wi-Fi probe requests}
\author{
	\authorstyle{
		Balamurugan Soundararaj\textsuperscript{1}, 
		James Cheshire\textsuperscript{1} and 
		Paul Longley\textsuperscript{1}}
	\newline\newline
	\textsuperscript{1}\institution{
		Department of Geography, 
		University College London, 
		United Kingdom}
}
\date{\today}

@@

@@

	\section{Abstract}

	In the past decade Wi-Fi has emerged as the most
	commonly used technology in providing 
	internet access to mobile devices such as smartphones, tablets
	and laptops in public and private spaces.
	This has resulted in multiple Wi-Fi networks being available at
	almost every location in an urban environment.
	Traversing through this overlapping mesh of Wi-Fi networks,
	modern mobile devices with Wi-Fi antennae regularly
	broadcast special type of signals known as probe requests to
	discover available Wi-Fi networks and switch seamlessly
	between them.
	This is a hardware level signal which was standardised by IEEE 802.11b/g
	specification relays information about the source mobile device to
	any Access Points (AP) available around it.
	Since this is the first step in establishing
	a connection between any two devices, it is
	is universal to any device which uses a Wi-Fi radio to communicate.
	This makes these probe requests
	an open, passive, continuous, and wireless source of
	data available at any urban location which can provide us with clues in
	understanding the number of people present in the immediate surroundings
	in real-time with high granularity \citep{freud2015,konto2017}.
	In this paper, from a set of probe requests collected
	at a high street location in London,
	along with manually counted data, we demonstrate
	that pedestrian footfall can be estimated with considerable accuracy without
	infringing on the privacy of the mobile users involved.

	\begin{figure*}
		\begin{center}
			\includegraphics[width=0.8\textwidth]{outputs/clustering_1.png}
			\caption{Clustering probe requests based on increasing sequence
			numbers present in them.}
		\end{center}
	\end{figure*}

	There have been numerous attempts at using Wi-Fi
	to measure the volume and movement of people in built environment for various
	applications \citep{zarim2006,sap2015,reki2007}.
	Though most research observe feasible and favorable results,
	in recent years, one of the major challenges faced in such attempts
	has been the MAC address randomisation process.
	This process aims to protect the users' privacy by anonymising
	the only globally identifiable portion of the probe requests
	resulting in a set of probe requests generated by the same device
	with different random MAC addresses \citep{green2008}.
	There have been various successful research in breaking this
	randomisation process to extract real MAC addresses \citep{martin2017}
	but this usually results in serious risk in infringement of
	privacy of the users of the mobile devices.
	There is a clear gap in the research for exploring methodologies
	which enables us to estimate the number of unique mobile
	devices from a set of anonymised probe requests
	without the need to reveal their original MAC addresses.

	A pilot survey was conducted at Oxford street in London
	on 20 December 2017 from 12:30 to 13:00 where two sets of data
	were collected on pedestrian footfall through Wi-Fi sensing
	and manual counting in parallel.
	The Wi-Fi sensor collected all the probe requests that were 
	broadcast around the area and recorded
	the time-stamp at which they were collected,
	MAC address of the source mobile device
	(anonymised using a hashing algorithm),
	organisationally unique identifier (OUI) of the manufacturer of the device,
	total length of the signal in bits,
	the strength of the signal reported by the mobile device in dBm,
	the sequence number of the signal,
	duration for which the signal was transmitted, 
	the service set identifier (SSID) of the access point targeted
	by the probe request and 
	the length of the extra information (tags) embedded in the packets.
	The manual count was done by an android application on a mobile phone
	which records the time-stamp every time the surveyor touches the screen
	which corresponds to one pedestrian footfall on the sidewalk.

	Being one of the busiest location in the United Kingdom, the location
	generated approximately 60,000 probe requests in the 30 minutes
	interval through the Wi-Fi sensor and 3,722 people were counted manually.
	When we aggregate the probe requests by their MAC address for every minute,
	the difference between the sensor counts and the manual counts
	is observed to be on average 425\%. 
	This shows that there is a large amount of noise
	in the data which might include signals from devices outside the area where
	the manual count was conducted and the anonymised probe requests from the
	same devices with different MAC addresses.
	Before we go look at the data
	in detail, an initial analysis shows that the fields - SSID and tags are
	very sparse and doesn't provide much information for our cleaning process 
	and the duration field is closely related to the length of the 
	probe request and provides no new information.
	So we remove these fields from further analysis.

	We remove the noise from devices outside the area of interest 
	by removing all the probe requests which report a "low" signal strength.
	This classification of "high" vs "low" is done through "k-means"
	classification algorithm.
	The cutoff point for the collected data is -71dBm.
	This process of filtering is highly effective and reduces the difference
	between the sensor counts and manual counts to 30\%. We observe that
	around 55\% of all probe requests collected are anonymised.
	We assign the hashed MAC address the unique identifier for the rest of
	the 45\% and investigate the anonymised probe requests further.
